---
title: "Homework 2"
author: "Steven Barnett"
date: "9/10/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 3

# a) Sensory Data

# First, I am going to pull down the dataset and save it to local storage
# Remote data: http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/Sensory.dat
```{r sensory_data_save_to_local, include=TRUE, eval=TRUE, echo=TRUE}
## sensory_data_url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/Sensory.dat"
## sensory_data <- fread(sensory_data_url, skip = 1, fill = TRUE, data.table = FALSE)
## saveRDS(sensory_data, "dwnldd_data/sensory_data_raw.RDS")
sensory_data <- readRDS("dwnldd_data/sensory_data_raw.RDS")
```

# The dataset has some incorrect row lengths due to indices being included in the data. We will fix these using Base R functions.
```{r sensory_data_base_r_clean, include=TRUE, eval=TRUE, echo=TRUE}
sensory_data_base_r <- sensory_data
for (i in seq(from = 1, to = 30, by = 3)) {
  sensory_data_base_r[i,] = sensory_data_base_r[i,][2:6]
}
sensory_data_base_r <- subset(sensory_data_base_r, select = -c(6))
names(sensory_data_base_r) <- c("op1", "op2", "op3", "op4", "op5")
```

# We will attempt to clean the dataset using tidyverse functions
```{r sensory_data_tidyverse_clear, include=TRUE, eval=TRUE, echo=TRUE}
#sensory_data_tidyverse <- sensory_data
#sensory_data_tidyverse <- select(sensory_data_tidyverse, -6)
#rename(sensory_data_tidyverse, op1 = Item, op2 = 1, op3 = 2, op4 = 3, op5 = 4)
```
# After cleaning the dataset using Base R and tidyverse functions, I am going to display the cleaned data
```{r sensory_data_display, include=TRUE, eval=TRUE, echo=FALSE}
knitr::kable(summary(sensory_data_base_r))
boxplot(sensory_data_base_r)
```


# b) Gold Medal Data

## First, I will pull the data down and store it locally

```{r gold_medal_data_save_to_local, include=TRUE, eval=TRUE, echo=TRUE}
## gold_medal_data_url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/LongJumpData.dat"
## gold_medal_data <- fread(gold_medal_data_url, fill=TRUE)
## saveRDS(gold_medal_data, "dwnldd_data/gold_medal_data_raw.RDS")
gold_medal_data <- readRDS("dwnldd_data/gold_medal_data_raw.RDS")
```

# The dataset has some incorrect row lengths due to indices being included in the data. We will fix these using Base R functions.
```{r gold_medal_data_clean_base_r, include=TRUE, eval=TRUE, echo=TRUE}
gold_medal_data_base_r <- gold_medal_data
gold_medal_data_frame <- data.frame(Year=integer(), Distance=numeric())
for (i in seq(from = 1, to = 11, by = 2)) {
  for (j in seq(from = 1, to = 6, by = 1)) {
    data_row <- gold_medal_data_base_r[j,]
    year <- data_row[[i]]
    distance <- data_row[[i+1]]
    gold_medal_data_frame <- rbind(gold_medal_data_frame, c(year, distance))
  }
}
names(gold_medal_data_frame) <- c("Year", "Distance")
gold_medal_data_frame <- na.omit(gold_medal_data_frame)
gold_medal_data_frame$Year <- gold_medal_data_frame$Year + 1900
```


# After cleaning the dataset using Base R and tidyverse functions, I am going to display the cleaned data
```{r gold_medal_data_display, include=TRUE, eval=TRUE, echo=FALSE}
knitr::kable(summary(gold_medal_data_frame))
plot(gold_medal_data_frame$Year, gold_medal_data_frame$Distance)
```
